\textsc{PAPERCLIP}: Associating Astronomical Observations and Natural Language with Multi-Modal Models

### 2024-03-11 paper/hubble_paperclip.tex
Started adding related works

The CLIP family of foundation models, which in their original form embed images and associated captions into a common representation space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval.
%
The concept of associating diverse modalities via contrastive training has been employed in other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations. Here, we present for the first time an application associating astronomical data with the text modality.

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=0.99\textwidth]{plots/figure.pdf}
\caption{\changes{Overview of the PAPERCLIP method. (Left) A pre-trained CLIP model is fine-tuned using a dataset of \hubble observations and corresponding proposal abstracts. (Right) The fine-tuned model can then be used for downstream tasks such as observation retrieval (i.e., finding the observations most relevant to a given text query). The proposal abstract snippet here corresponds to proposal ID \href{https://archive.stsci.edu/proposal_search.php?id=16914&mission=hst}{16914}}}
\label{fig:overview}
\end{figure*}

@@

\subsection{Contrastive Language-Image Pre-training (CLIP)}

CLIP \citep[Contrastive Language-Image Pre-training;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
We use the standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \label{eq:softmax_loss}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are considered on the same footing.

@@

We also qualitatively evaluate the learned embeddings through image retrieval (i.e., retrieving the most relevant images from the validation set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation, akin to zero-shot classification) experiments. 
%
For the description/text retrieval evaluation, we define a list of possible text associations (i.e., classes), which we show in App.~\ref{app:categories}, by querying the \textsc{Claude 2}\footnote{\url{https://claude.ai/}} large language model along with manual curation.

>>>>

\paragraph*{Related work}

% The CLIP family of foundation models, which in their original form embed images and associated captions into a common representation space via contrastive learning, have shown strong performance and generalization capabilities on a variety of downstream tasks including zero-shot classification and image retrieval~\citep{radford2021learning}.
%
The concept of learning task-agnostic representations via self-supervised learning has been applied within astrophysics \cite{slijepcevic2024radio,stein2021self,hayat2021self,slijepcevic2022learning} and used for downstream tasks like object similar search \citep{stein2021self}, gravitational lens finding \citep{stein2022mining}, estimation of Galactic distances \citep{hayat2021estimating}, and identification of rare galaxies \citep{walmsley2023rare}.
%
Associating different modalities via contrastive training has been employed in many other scientific domains~\citep[e.g.,][]{liu2023text,Sanchez-Fernandez2022.11.17.516915,lanusse2023astroclip,cepeda2023geoclip}, and has been shown to be effective in learning semantically meaningful joint representations. Here, we present for the first time an application associating diverse astronomical data with the text modality.
%
For a recent review of contrastive learning in astrophysics, see \citet{huertas2023brief}. 
%
% \citet{bowles2023radio,bowles2022new}

@@

\begin{figure*}[!t]
\centering
\includegraphics[width=0.99\textwidth]{plots/figure.pdf}
\caption{\changes{Overview of the PAPERCLIP method. (Left) A pre-trained CLIP model is fine-tuned using a dataset of \hubble observations and corresponding proposal abstracts. (Right) The fine-tuned model can then be used for downstream tasks such as observation retrieval (i.e., finding the observations most relevant to a given text query). The proposal abstract snippet here corresponds to proposal ID \href{https://archive.stsci.edu/proposal_search.php?id=16914&mission=hst}{16914}}.}
\label{fig:overview}
\end{figure*}

@@

\subsection{Contrastive Language-Image Pre-training}

Contrastive Language-Image Pre-training \citep[CLIP;][]{radford2021learning} is a multi-modal neural network model pre-trained on a large corpus of image-text pairs via weak supervision using a contrastive loss.
%
Given a minibatch $\mathcal{B}$ of $|\mathcal{B}|$ image-text pairs $\{(I_i, T_i)\}$, the goal is to align the learned representations of corresponding (positive) pairs $(I_i, T_i)$ while repelling the representations of unaligned (negative) pairs $(I_i, T_{j\neq i})$.
%
Image and text encoders $f: I \rightarrow \mathbb R^{n_\text{emb}}$ and $g: T \rightarrow \mathbb R^{n_\text{emb}}$ are used to map images and text to a common embedding space of dimension $n_\text{emb}$.
%
We use the standard softmax-based bidirectional variant of the InfoNCE~\citep{oord2018representation} contrastive loss function introduced for training CLIP-style architectures \citep{radford2021learning}
%
\begin{equation}
  \label{eq:softmax_loss}
  \mathcal{L}(\mathcal{B})=-\frac{1}{2|\mathcal{B}|} \sum_{i=1}^{|\mathcal{B}|}\left(\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_i \cdot y_j / \tau}}+\log \frac{e^{x_i \cdot y_i / \tau}}{\sum_{j=1}^{|\mathcal{B}|} e^{x_j \cdot y_i / \tau}}\right)
\end{equation}
%
where ${x}_i={f\left(I_i\right)}/{\left\|f\left(I_i\right)\right\|}$ and ${y}_i={g\left(T_i\right)}/{\left\|g\left(T_i\right)\right\|}$ are the normalized representations of the $i$-th image and text, respectively, and $\tau$ is a learnable temperature hyperparameter.
%
Note that this loss treats the image and text representations symmetrically, ensuring that the two modalities are considered on the same footing.

@@

We also qualitatively evaluate the learned embeddings through image retrieval (i.e., retrieving the most relevant images from the validation set using natural language queries) and description retrieval (i.e., querying the astrophysical object classes and science use cases most relevant to a given observation, akin to zero-shot classification) experiments. 
%
For the description/text retrieval evaluation, we define a list of possible text associations (i.e., classes), which we show in App.~\ref{app:categories}, by querying the \textsc{Claude 2}\footnote{\url{https://claude.ai/}} large language followed by with manual curation.
